---
title: "Final Project Preiminary EDA"
author: "Josiah Kim"
date: "13 November 2020"
output: html_notebook
---

### Guiding Question
Do black students face harsher/more punishment than their counterparts within the public school system?  


### Become Acquainted with the Data
I found the data on The Civil Rights Project (https://www.civilrightsproject.ucla.edu/) who aim to renew the civil rights movement by creating a deeper understanding of the disparities. This data is collected and maintained by The Civil Rights Data Collection (https://ocrdata.ed.gov/). Every other year, all public schools are required to submit a reponse to a survey to the U.S. Department of Education's Office for Civil Rights. I will be using *Referrals and Arrests* dataset which shows a record of students' referrals and law enforcement interention. Here is a peek at the data:
```{r}
# Load necessary packages
library(tidyverse)
# Stage the data
refAndArrest <- read.csv("Referrals and Arrests.csv")
# Start data exploration 
glimpse(refAndArrest)
```
There are 97, 632 cases in this dataset containing 84 variables. Each case represents a public school. I plan to use the following variables:

- *SCH_DISCWODIS_REF_HI_M* - Students without disabilities who were referred to a law enforcement agency or official: Hispanic Male
- *SCH_DISCWODIS_REF_HI_F* - Students without disabilities who were referred to a law enforcement agency or official: Hispanic Female
- *SCH_DISCWODIS_REF_AM_M *- Students without disabilities who were referred to a law enforcement agency or official: American Indian/Alaska Native Male
- *SCH_DISCWODIS_REF_AM_F* - Students without disabilities who were referred to a law enforcement agency or official: American Indian/Alaska Native Female
- *SCH_DISCWODIS_REF_AS_M* - Students without disabilities who were referred to a law enforcement agency or official: Asian Male
- *SCH_DISCWODIS_REF_AS_F* - Students without disabilities who were referred to a law enforcement agency or official: Asian Female
- *SCH_DISCWODIS_REF_HP_M* - Students without disabilities who were referred to a law enforcement agency or official: Native Hawaiian/Pacific Islander Male
- *SCH_DISCWODIS_REF_HP_F* - Students without disabilities who were referred to a law enforcement agency or official: Native Hawaiian/Pacific Islander Female
- *SCH_DISCWODIS_REF_BL_M* - Students without disabilities who were referred to a law enforcement agency or official: Black Male
- *SCH_DISCWODIS_REF_BL_F* - Students without disabilities who were referred to a law enforcement agency or official: Black Female
- *SCH_DISCWODIS_REF_WH_M* - Students without disabilities who were referred to a law enforcement agency or official: White Male
- *SCH_DISCWODIS_REF_WH_F* - Students without disabilities who were referred to a law enforcement agency or official: White Female
- *SCH_DISCWODIS_REF_TR_M* - Students without disabilities who were referred to a law enforcement agency or official: Two or More Races Male
- *SCH_DISCWODIS_REF_TR_F* - Students without disabilities who were referred to a law enforcement agency or official: Two or More Races Female
- *SCH_DISCWODIS_ARR_HI_M* - Students without disabilities who received a school-related arrest: Hispanic Male
- *SCH_DISCWODIS_ARR_HI_F* - Students without disabilities who received a school-related arrest: Hispanic Female
- *SCH_DISCWODIS_ARR_AM_M* - Students without disabilities who received a school-related arrest: American Indian/Alaska Native Male
- *SCH_DISCWODIS_ARR_AM_F* - Students without disabilities who received a school-related arrest: American Indian/Alaska Native Female
- *SCH_DISCWODIS_ARR_AS_M* - Students without disabilities who received a school-related arrest: Asian Male
- *SCH_DISCWODIS_ARR_AS_F* - Students without disabilities who received a school-related arrest: Asian Female
- *SCH_DISCWODIS_ARR_HP_M* - Students without disabilities who received a school-related arrest: Native Hawaiian/Pacific Islander Male
- *SCH_DISCWODIS_ARR_HP_F* - Students without disabilities who received a school-related arrest: Native Hawaiian/Pacific Islander Female
- *SCH_DISCWODIS_ARR_BL_M* - Students without disabilities who received a school-related arrest: Black Male
- *SCH_DISCWODIS_ARR_BL_F* - Students without disabilities who received a school-related arrest: Black Female
- *SCH_DISCWODIS_ARR_WH_M* - Students without disabilities who received a school-related arrest: White Male
- *SCH_DISCWODIS_ARR_WH_F* - Students without disabilities who received a school-related arrest: White Female
- *SCH_DISCWODIS_ARR_TR_M* - Students without disabilities who received a school-related arrest: Two or More Races Male
- *SCH_DISCWODIS_ARR_TR_F* - Students without disabilities who received a school-related arrest: Two or More Races Female

### Explore Intuition
First, I will create a new data frame containing just the variables that I need.
```{r}
cleanedrefAndArrest
```


### TWO Data Sources
